---
title: "DASI_Project"
author: "Todd Robinson"
date: "Tuesday, October 14, 2014"
output: html_document
---
---

<!-- For more info on RMarkdown see http://rmarkdown.rstudio.com/ -->

<!-- Enter the code required to load your data in the space below. The data will be loaded but the line of code won't show up in your write up (echo=FALSE) in order to save space-->
```{r echo=FALSE}
rm(list=ls())

#Load some required packages
require(Amelia) # missmap needs this
require(Hmisc) # bystats
load(url("http://bit.ly/dasi_gss_data"))

```

<!-- In the remainder of the document, add R code chunks as needed -->

### Introduction:

The U.S. subprime mortgage crisis was a nationwide banking emergency that coincided with the U.S. recession of December 2007-June 2009.[1]
There were many causes of the crisis, with commentators assigning different levels of blame to financial institutions, regulators, credit agencies, government housing policies, and consumers, among others.[2] 
Is there a relationship between the working Respondent's feelings about losing their job to their confidence in U.S. Financial Institutions?
Did the problems with the U.S Economy in 2007, which were largely due to U.S. Financial Institutions, have an effect on how Worker's feel about the possibility of losing their job?
In the past, job security was a function of how a Worker felt about the economy. That may have changed to revolve around, not the economy, rather the U.S. Finanacial Instutions.

### Data:

The vast majority of GSS data is obtained in face-to-face interviews. Computer-assisted personal interviewing (CAPI) began in the 2002 GSS. Under some conditions when it has proved difficult to arrange an in-person interview with a sampled respondent, GSS interviews may be conducted by telephone. The GSS was first conducted in 1972. Until 1994, it was conducted almost annually (due to funding limitations, there were no GSSs in 1979, 1981, or 1992). Since 1994, the GSS has been conducted in even numbered years. New GSSs are planned for 2008 and 2010.
Each case is a response to a GSS survey. Each consists of how a person responding to the survey answers each of the question on his or her questionnaire. Each row or case is a survey response while the columns or variables are the answers to each question on the questionnaire (survey). A subset of these observations was taken to limit this analysis to just those Respondents who were currently working at the time of their survey responses. 
gss$wrkstat = 
Working Fulltime
Working Parttime
Temp

This is an observational study. It is a retrospective observational study seeking to determine if there is a relationship between a person's age and how secure they feel in the job. It is not an experiment. This study is not changing or "experimenting" with the data to determine relationships within the data. This study will not be randomizing cases into treatment and control groups in order to try different experiments on the treatment group to see if they have a effect , correlation or other relationship between the experiment and outcome. We are observing the data already collected and will be analyzing that data to answer a specific question to see if there is a relationship between two data elements.

### Exploratory Data Analysis:
####Step 1 - Subset the gss data to only those rows where Respondents are in (or have been) in the workforce.
```{r echo=TRUE}
#Subset the data
workers <- c("Working Fulltime", "Working Parttime", "Temp Not Working", "Unempl, Laid Off")
df.gss<-subset(gss,wrkstat %in% workers)

df.gss$wrkstat<-factor(df.gss$wrkstat)
```
####Step 2 - Subset the features in the gss data to only those that might be relevant to this Analysis.
```{r echo=TRUE}
fields<-c("year","age","sex","wrkstat","confinan","conbus","joblose","satfin","finrela") #jobsec too sparse. Looks like parsol wasn't asked until after 1993
df.gss<-df.gss[,fields]
dim(df.gss)
```

####Step 3 - Massage data to get rid of NAs.
Take a look at the data overall. Specifically missing (NA) data in the principal features of interest confinan and joblose

```{r echo=FALSE}
#Check missing values in the data

par(mar=c(5,1,5,2))
par(mfrow=c(1,2))
main<-"Figure 1\nBEFORE\ngss data"
col<-c("yellow", "black")
missmap(df.gss,main=main,col=col,legend=FALSE,y.cex=0.001,x.cex=1.0)
df.gss<-subset(df.gss,!is.na(df.gss$confinan) & !is.na(df.gss$joblose),select=fields)
main<-"Figure 2\nAFTER\nFiltering out NAs"
col<-c("yellow", "black")
missmap(df.gss,main=main,col=col,legend=FALSE,y.cex=0.001,x.cex=1.0)

```

Figure 1:
Before any filtering. Lots of missing data in the features of primary interest.
The dataset has 57,061 rows. Let's subset out the missing data from the primary features (confinan and joblose).

Figure 2: 
After filtering out rows where features confinan AND joblose contain NA. This leaves us with a dataset of 10,978.
It is interesting to note that confinan and joblose NAs appear to be related in some way because removing where they BOTH are NAs
has removed ALL NAs from either one. Looks promising. Let's continue with some inference on this dataset.

####Step 4 - Is there a difference between Confidence in Financial Insitutions and Job security in the 2 years leading up to 2007 compared to the two years after 2007?

```{r echo=FALSE}
df.gss06<-df.gss[df.gss$year %in% c(2004,2006),]
df.gss08<-df.gss[df.gss$year > 2008,]

par(mfrow=c(1,2))
plot(df.gss06$confinan~df.gss06$joblose);plot(df.gss08$confinan~df.gss08$joblose)


```

### Inference:

Checking for Independence
1. The gss dataset 
### Conclusion:

Insert conclusion here...

### References:
####Data URL
http://bit.ly/dasi_gss_data 

####Data Citation


[2]Smith, Tom W., Michael Hout, and Peter V. Marsden. General Social Survey, 1972-2012 [Cumulative File]. ICPSR34802-v1. Storrs, CT: Roper Center for Public Opinion Research, University of Connecticut /Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributors], 2013-09-11. doi:10.3886/ICPSR34802.v1

Persistent URL: http://doi.org/10.3886/ICPSR34802.v1

####Other Citations

[1] US Business Cycle Expansions and Contractions, NBER, accessed August 9, 2012. http://www.nber.org/cycles.html

[2]Financial Inquiry Commission-Final Report-Retrieved February 2013.  http://fcic.law.stanford.edu/report
